---
title: "Structural Bioinformatics Class 12"
author: "Ashley Lee"
date: "5/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1: In Silico Docking of Drugs to HIV-1 Protease 

## 1.1 Obtaining and Inspecting Our Input Structure 

Load Bio3D package and obtain PDB file of "1HSG" from RCSB PDB database.
```{r}
library(bio3d)
file.name <- get.pdb("1hsg")
```

Read the PDB structure into R. 
```{r}
hiv <- read.pdb(file.name)
```

Quick summary of "hiv."
```{r}
hiv
```

## 1.2 Prepare Initial Protein and Ligand Input Files 

Separate new protein only PDB format file and separate ligand only PDB format file. 
```{r}
prot <- trim.pdb(hiv, "protein")
lig <- trim.pdb(hiv, "ligand")
```

Write out these new PDB files so we can use them for docking
```{r}
write.pdb(prot, file="1hsg_protein.pdb")
write.pdb(lig, file="1hsg_ligand.pdb")
```

## 1.3 Using AutoDockTools to Setup Protein Docking Input
Download AutoDockTools and setup protein

## 1.4 Prepare the Ligand 
Prepare ligand using ADT

## 1.5 Prepare a Docking Configuration File 
Create "config.txt" file 

# Section 2: Docking Ligands into HIV-1 Protease

## 2.1 Download Autodock Vina 
Download Autodock Vina and access through terminal

## 2.2 Docking Indinavir into HIV-1 protease  
Using terminal, dock Indinvair into HIV-1 protease and generate "all.pdbqt""

## 2.3 Inspecting Your Docking Results 

```{r}
library(bio3d)
res <- read.pdb("all.pdbqt", multi=TRUE)

write.pdb(res, "result.pdb")
```

```{r}
ori <- read.pdb("1hsg_ligand.pdbqt")

rmsd(ori, res)
```

```{r}
inds.res <- atom.select(res,"noh")
```

```{r}
# res$xyz[,inds.res$xyz]
rmsd(lig, res$xyz[,inds.res$xyz])
```

# OPTIONAL: Section 3: Exploring the Conformational Dynamics of Proteins 

Watch introduction video.






















