---
title: "Lecture11"
author: "Ashley Lee"
date: "5/8/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1: Introduction to the RCSB Protein Data Bank (PDB)
**Q1.** What proportion of PDB entries does X-ray crystallography account for? What proportion of
structures are protein?
**A1.** Yearly: 3135; Total: 125271

```{r}
pdb.stats <- read.csv("Data Export Summary.csv")
```

Let's calculate something!

```{r}
percent <- (pdb.stats$Total / sum (pdb.stats$Total)*100)
names(percent) <- pdb.stats$Experimental.Method
percent 
```

**Q2.** Type HIV in the search box on the home page and determine how many HIV-1 protease
structures are in the current PDB?


## Section 2: Visualizing the HIV-1 Protease Structure
In this section we will use the 2Å resolution X-ray crystal structure of HIV-1 protease with a bound drug molecule indinavir (PDB ID: 1HSG) [3]. We will use the VMD molecular viewer to
visually inspect the protein, the binding site and the drug molecule. 

**Q3.** Water molecules normally have 3 atoms. Why do we see just one atom per water molecule
in this structure?

**Q4.** There is a conserved water molecule in the binding site. Can you identify this water
molecule? What residue number does this water molecule have?

**Q5.** As you have hopefully observed HIV protease is a homodimer (i.e. it is composed of two identical chains). With the aid of the graphic display and the sequence viewer extension can you identify secondary structure elements that are likely to only form in the dimer rather than the monomer?


## Section 3: Introduction to Bio3D in R

### 3.1: Getting Started 

```{r}
library(bio3d)
```

### 3.2: Bio3D Functions and Their Typical Usage

```{r}
example(plot.bio3d)
```

### 3.3 Working with Individual PDB Files

### 3.4 Reading PDB file data into R

```{r}
pdb <- read.pdb("1hsg")
print(pdb)
```

**Q6.** How many amino acid residues are there in this pdb object and what are the two nonprotein residues?
**A6.** There are 198 residues and the two nonprotein residues are water and MK1.

```{r}
# To find the attributes of any such object
attributes(pdb)
```

```{r}
# Access the atom attribute of "pdb""
head(pdb$atom)

# Print a subset of $atom data for the first two atoms
pdb$atom[1:2, c("eleno", "elety", "x","y","z")]

# Note that individual $atom records can also be accessed like this
pdb$atom$elety[1:2]

# Which allows us to do the following
plot.bio3d(pdb$atom$b[pdb$calpha], sse=pdb, typ="l", ylab="B-factor")
```

**Q7.** What type of R object is pdb$atom? HINT: You can always use the str() function to get a useful summery of any R object.

```{r}
# Print a summary of the coordinate data in $xyz
pdb$xyz

# Examine the row and column dimensions
dim(pdb$xyz)

# Print coordinates for the first two atom
pdb$xyz[ 1, atom2xyz(1:2) ]
```

### 3.5 Atom selection
```{r}
# Select all C-alpha atoms (return their indices)
## ca.inds <- atom.select(pdb, "calpha")
## ca.inds

# Print details of the first few selected atoms
## head( pdb$atom[ca.inds$atom, ]

# And selected xyz coordinates
## head( pdb$xyz[, ca.inds$xyz] )
```

```{r}
# Select chain A
## a.inds <- atom.select(pdb, chain="A")

# Select C-alphas of chain A
## ca.inds <- atom.select(pdb, "calpha", chain="A")

# We can combine multiple selection criteria to return their intersection
## cab.inds <- atom.select(pdb, elety=c("CA","CB"), chain="A", resno=10:20)
```

**Q8.** Use the Bio3D write.pdb() function to write out a new C-alpha atom only PDB file for
viewing in VMD.
**A8.**

```{r}
# Select all C-alpha atoms (return their indices)
inds <- atom.select(pdb, "calpha")
inds
```

```{r}
# Check if we have what we want.
head(pdb$atom[inds$atom,])
```

```{r}
# Looks good, so let's continue to trim our PDB and write it out
ca.pdb <- trim.pdb(pdb, inds=inds)
ca.pdb

write.pdb(ca.pdb, file="calpha_only.pdb")
```

Open resulting file in VMD!

## Section 4: Working with Multiple PDB Files

### 4.1 Installing the Stand-Alone Muscle Alignment Program
Download MUSCLE program! 

### 4.2 Aligning Multiple Structures
```{r}
# Download some example PDB files
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)
```

```{r}
# Extract and align the chains we are interested in
pdbs <- pdbaln(files, fit = TRUE)

# Print to screen a summary of the 'pdbs' object
pdbs
```

**Q9.** What effect does setting the fit=TRUE option have in the related rmsd() function? What does RMSD measure and what would the results indicate if you set fit=FALSE or
removed this option?

```{r}
# Access the first 5 rows, and 8 columns
pdbs$ali[1:5, 1:8]

# Associated residues numbers
pdbs$resno[1:5, 1:8]
```

```{r}
# Inspect alignment for gaps
gaps <- gap.inspect(pdbs$ali)

# Indices of non-gap columns
head(gaps$f.inds)

# Access non-gap columns
pdbs$ali[, gaps$f.inds]
```

### Section 4.3 Basic structure analysis

```{r}
# Calculate sequence identity
seqidentity(pdbs)

# Calculate RMSD
rmsd(pdbs)
```

## Section 5: Example Application on Adenylate Kinase (Adk)

### 5.1 Search and Retrieve Adenylate Kinase Structures

```{r}
aa <- get.seq("1ake_A")
aa
```

```{r, cache=TRUE}
# Blast or hmmer search
b <- blast.pdb(aa)
```

```{r}
# Plot a summary of search results
hits <- plot(b)
```

```{r}
head(hits$pdb.id)
```

```{r}
# Fetch PDBs
files <- get.pdb(hits$pdb.id, path = "pdbs", split = TRUE, gzip = TRUE)

# Align structures
pdbs <- pdbaln(files)

# Vector containing PDB codes
ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
plot(pdbs, labels=ids)
```


### Section 5.2: Sequence Conservation Analysis

```{r}
# Calculate sequence conservation
cons <- conserv(pdbs, method="entropy22")

# SSE annotations
sse <- pdbs2sse(pdbs, ind=1, rm.gaps=FALSE)

# Plot conservation per residue
plotb3(cons, sse=sse, ylab="Sequence entropy")
```

### Section 5.3: Annotate collected PDB structures

```{r}
# anno <- pdb.annotate(ids) - Need to install XML and RCurl package from CRAN
# print(unique(anno$source))
```

### Section 5.4: Principal Component Analysis 

```{r}
# Find invariant core
core <- core.find(pdbs)

# Superimpose all structures to core
pdbs$xyz = pdbfit(pdbs, core)

# Perform PCA
pc.xray <- pca(pdbs)
```

```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

```{r}
# Left-click on a point to label and right-click to end
# identify (pc.xray$z[,1:2], labels=basename.pdb(pdbs$id)) - error
```




















